Two subtle issues with over-representation analysis: Demonstration

### 2024-07-01 app/report.Rmd
progress on report

---
title: "Two subtle issues with over-representation analysis: Demonstration"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
params:
  fg: NA
  bg: NA
  comparison: NA
  genesetlibrary: NA
---

@@

## Genes without annotated categories are discarded

@@

## p-value adjustment is applied incorrectly

@@

Some information about the parameters used:

@@

fg <- params$fg
bg <- params$bg
comparison <- params$comparison
genesetlibrary <- params$genesetlibrary
minsetsize <- 5

message(paste("Minimum gene set size:",minsetsize))

@@

Here is the first few lines of the input foreground and background lists.

@@

```{r,peek}

message("First 10 genes in the foreground")
head(fg,10)

message("First 10 genes in the background")
head(bg,10)

message("Num genes in foreground")
length(fg)

message("Num genes in background")
length(bg)

@@

@@

Overlap of genes in the gene set library.

```{r,gsl2}

print("abc")

@@

print("run clusterprofiler default")

print("allow download?")

@@

Make bar chart.

@@

barplot(c(1,2,3))

@@

mymessage <- paste("Here we are characterising","comparison")
message(mymessage)

print("run clusterprofiler fixed")
# actually there will be three different types of analysis possible here
# depending on the selection

print("allow download?")

@@

Make bar chart.

@@

barplot(c(1,2,3))

@@

* Data table

* Euler diagram

* Scatterplot of ES

* Scatterplot of logFDR

>>>>

---
title: "Two subtle issues with over-representation analysis: Demonstration"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
params:
  fg: NA
  bg: NA
  comparison: NA
  genesetlibrary: NA
  mygs: NA
---

@@

### Genes without annotated categories are discarded

@@

### p-value adjustment is applied incorrectly

@@

@@

fg <- params$fg
bg <- params$bg
comparison <- params$comparison
genesetlibrary <- params$genesetlibrary
mygs <- params$mygs
minsetsize = 5

@@

@@

### Notes

First 10 genes in the foreground.

```{r,peek1}
head(fg,10)
```

First 10 genes in the background.

```{r,peek2}
head(bg,10)
```

Number of genes in the foreground list.

```{r,peek3}
length(fg)
```

Number of genes in the background list.

```{r,peek4}
length(bg)
```

## Gene set library

@@

nsets <- length(unique(mygs$term))
message(paste("Number of sets:",nsets))
```

Summary of gene set length.

```{r,gsl2}
summary(table(mygs$term))
```

Total number of genes with one or more annotations in the library.

```{r,gsl3}
length(unique(mygs$gene))
```

Total number of annotations in the gene set library.

```{r,gsl4}
nrow(mygs)
```

Overlap of foreground genes in the gene set library.

```{r,gsl5}

mygs_genes <- unique(mygs$gene)

I <- length(intersect(fg,mygs_genes))

message(paste(I,"from",length(fg)))

@@

Overlap of background genes in the gene set library.

```{r,gsl6}

I <- length(intersect(bg,mygs_genes))

message(paste(I,"from",length(bg)))

@@

options(enrichment_force_universe = FALSE)
ora <- as.data.frame(enricher(gene = fg ,
                              universe = bg, minGSSize = 5, maxGSSize = 500000, TERM2GENE = mygs,
                              pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

gr <- as.numeric(sapply(strsplit(ora$GeneRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora$GeneRatio,"/"),"[[",2))

br <- as.numeric(sapply(strsplit(ora$BgRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora$BgRatio,"/"),"[[",2))

ora$ES <- gr/br
ora$ID = ora$geneID = ora$p.adjust = ora$Count = NULL
colnames(ora) <- gsub("qvalue","FDR",colnames(ora))
colnames(ora) <- gsub("GeneRatio","FgRatio",colnames(ora))
ora <- ora[,c("Description","FgRatio","BgRatio","ES","pvalue","FDR")]
ora$ES <- signif(ora$ES,3)
ora$pvalue <- signif(ora$pvalue,3)
ora$FDR <- signif(ora$FDR,3)
ora1 <- ora

@@

```{r,defaulttable}

DT::datatable(
  { ora1 },

  extensions = 'Buttons',
  filter = list(position = 'top', clear = FALSE),

  options = list(
    paging = TRUE,
    searching = TRUE,
    fixedColumns = TRUE,
    autoWidth = TRUE,
    ordering = TRUE,
    dom = 'tB',
    buttons = c('copy', 'csv', 'excel'),
    pageLength = 50,
    search = list(regex = TRUE, caseInsensitive = TRUE)),

  rownames= FALSE)

Bar chart.

@@

edata <- head(subset(ora1,FDR<0.05),20)
if (nrow(edata)>1) {
  edata <- edata[order(edata$ES),]
  vec <- edata$ES
  names(vec) <- edata$Description
  names(vec) <- substr(names(vec), start = 1, stop = 60)
  names(vec) <- gsub("_"," ",names(vec))
  par(mar=c(4,27,1,1))
  barplot(vec,horiz=TRUE,las=1,cex.names=0.8,
          xlab="Enrichment Score")
  mtext("Default ORA",side=3,adj=1)
} else {
  message("No gene sets with FDR<0.05 found")
}

@@

mymessage <- paste("Here we are characterising",comparison)
message(mymessage)

#Background error

if (comparison == "Background error") {
  options(enrichment_force_universe = TRUE)
  ora <- as.data.frame(enricher(gene = fg ,
                                universe = bg, minGSSize = 5, maxGSSize = 500000, TERM2GENE = mygs,
                                pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

  gr <- as.numeric(sapply(strsplit(ora$GeneRatio,"/"),"[[",1)) /
    as.numeric(sapply(strsplit(ora$GeneRatio,"/"),"[[",2))

  br <- as.numeric(sapply(strsplit(ora$BgRatio,"/"),"[[",1)) /
    as.numeric(sapply(strsplit(ora$BgRatio,"/"),"[[",2))

  ora$ES <- gr/br
  ora$ID = ora$geneID = ora$p.adjust = ora$Count = NULL
  colnames(ora) <- gsub("qvalue","FDR",colnames(ora))
  colnames(ora) <- gsub("GeneRatio","FgRatio",colnames(ora))
  ora <- ora[,c("Description","FgRatio","BgRatio","ES","pvalue","FDR")]
  ora$ES <- signif(ora$ES,3)
  ora$pvalue <- signif(ora$pvalue,3)
  ora$FDR <- signif(ora$FDR,3)
}

ora2 <- ora
```

```{r,fixed2}

DT::datatable(
  { ora2 },

  extensions = 'Buttons',
  filter = list(position = 'top', clear = FALSE),

  options = list(
    paging = TRUE,
    searching = TRUE,
    fixedColumns = TRUE,
    autoWidth = TRUE,
    ordering = TRUE,
    dom = 'tB',
    buttons = c('copy', 'csv', 'excel'),
    pageLength = 50,
    search = list(regex = TRUE, caseInsensitive = TRUE)),

  rownames= FALSE)

@@

Fixed bar chart.

@@

edata <- head(subset(ora2,FDR<0.05),20)
if (nrow(edata)>1) {
  edata <- edata[order(edata$ES),]
  vec <- edata$ES
  names(vec) <- edata$Description
  names(vec) <- substr(names(vec), start = 1, stop = 60)
  names(vec) <- gsub("_"," ",names(vec))
  par(mar=c(4,27,1,1))
  barplot(vec,horiz=TRUE,las=1,cex.names=0.8,
          xlab="Enrichment Score")
  mtext("Fixed ORA",side=3,adj=1)
} else {
  message("No gene sets with FDR<0.05 found")
}

@@

Side by side data table.
Columns labeled 'x' are original algorithm, and 'y' are fixed.

```{r,comparison1}
m <- merge(ora1,ora2,by="Description")
m <- m[,c("Description","FgRatio.x","FgRatio.y","BgRatio.x","BgRatio.y","ES.x","ES.y","FDR.x","FDR.y")]
diff <- abs(-log10(m$FDR.x) - -log10(m$FDR.y))
m <- m[order(-diff),]
m$ES.x <- signif(m$ES.x,3)
m$ES.y <- signif(m$ES.y,3)
m$FDR.x <- signif(m$FDR.x,3)
m$FDR.y <- signif(m$FDR.y,3)

DT::datatable( { m },
  extensions = 'Buttons',
  filter = list(position = 'top', clear = FALSE),
  options = list(
    paging = TRUE,
    searching = TRUE,
    fixedColumns = TRUE,
    autoWidth = TRUE,
    ordering = TRUE,
    dom = 'tB',
    buttons = c('copy', 'csv', 'excel'),
    pageLength = 50,
    search = list(regex = TRUE, caseInsensitive = TRUE)),
  rownames= FALSE)

Euler diagram.

```{r,comparison2}

ora1_sets <- subset(ora1,FDR < 0.05)$Description
ora2_sets <- subset(ora2,FDR < 0.05)$Description

if ( comparison == "Background error") {
  v1 <- list("Original"=ora1_sets, "BG fix"=ora2_sets)
}
plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2))

Scatterplot of ES.

```{r,comparison3}

orig_df <- ora1
bgfix_df <- ora2
m <- merge(orig_df,bgfix_df,by="Description")
m <- m[,c("Description","ES.x","ES.y")]
MAX <- max(c(m$ES.x,m$ES.y))
INCREMENT <- MAX/nrow(m)
VEC <- seq(0,MAX,INCREMENT)

if ( comparison == "Background error") {

m$Original <- m$BGcorrected <- round(VEC[1:nrow(m)],0.1)
m$ES.x <- signif(m$ES.x,3)
m$ES.y <- signif(m$ES.y,3)

plot_ly(
  m, x = ~ES.x, y = ~ES.y
) %>%
  add_trace(m, x = ~Original, y = ~BGcorrected, type = "scatter",
            mode="lines", line=list(color='grey')) %>%
  add_trace(m, x = ~ES.x, y = ~ES.y, type = "scatter", mode = "markers",
            showlegend = FALSE, text = m$Description,
            hoverinfo = 'text') %>%
  layout(title="Enrichment score comparison")
}

Scatterplot of logFDR values.

```{r,comparison4}

m <- merge(ora1,ora2,by="Description")
m <- m[,c("Description","FDR.x","FDR.y")]
m$logFDR.x <- -log10(m$FDR.x)
m$logFDR.y <- -log10(m$FDR.y)
MAX <- max(c(m$logFDR.x,m$logFDR.y))
INCREMENT <- MAX/nrow(m)
VEC <- seq(0,MAX,INCREMENT)

if ( comparison == "Background error") {

  m$Original <- m$BGcorrected <- round(VEC[1:nrow(m)],0.1)
  m$logFDR.x <- signif(m$logFDR.x,3)
  m$logFDR.y <- signif(m$logFDR.y,3)

  plot_ly(
    m, x = ~logFDR.x, y = ~logFDR.y
  ) %>%
    add_trace(m, x = ~Original, y = ~BGcorrected, type = "scatter",
              mode="lines", line=list(color='grey')) %>%
    add_trace(m, x = ~logFDR.x, y = ~logFDR.y, type = "scatter", mode = "markers",
              showlegend = FALSE, text = m$Description,
              hoverinfo = 'text') %>%
    layout(title="log FDR comparison")

}
